---
title: "Project4"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

In this project, you are going to explore matrix factorization methods for recommender system. The goal is to match consumers with most appropriate products. Matrix factorization methods characterize both items and users by vectors of factors inferred from item rating patterns. High correspondence between item and user factors leads to a recommendation. Matrix factorization generally has 3 parts:

- factorization algorithm

- regularization

- postpocessing

It is highly recommended to read this [review paper](./paper/P1 Recommender-Systems.pdf).

### Step 1 Load Data and Train-test Split
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
data <- read.csv("../data/ml-latest-small/ratings.csv")
set.seed(0)
test_idx <- sample(1:nrow(data), round(nrow(data)/5, 0))
train_idx <- setdiff(1:nrow(data), test_idx)
data_train <- data[train_idx,]
data_test <- data[test_idx,]
```

###Step 2 Matrix Factorization
#### Step 2.1 Algorithm and Regularization
Here I perform stochastic gradien descent to do matrix factorization.
Your algorithm should consider case that there are new users and movies adding to the dataset you used to train. In other words, the dimension your matrix R, q, p is dynamic.

- For algorithms, the referenced paper are:

A1. [Stochastic Gradient Descent](./paper/P1 Recommender-Systems.pdf) Section: Learning Algorithms-Stochastic Gradient Descent

A2. [Gradient Descent with Probabilistic Assumptions](./paper/P3 probabilistic-matrix-factorization.pdf) Section 2

A3. [Alternating Least Squares](./paper/P4 Large-scale Parallel Collaborative Filtering for the Netflix Prize.pdf) Section 3.1

- For regularizations, the referenced paper are:

R1. [Penalty of Magnitudes](./paper/P1 Recommender-Systems.pdf) Section: a Basic Matrix Factorization Model

R2. [Bias and Intercepts](./paper/P1 Recommender-Systems.pdf) Section: Adding Biases

R3. [Temporal Dynamics](./paper/P5 Collaborative Filtering with Temporal Dynamics.pdf) Section 4


```{r}
U <- length(unique(data$userId))
I <- length(unique(data$movieId))
source("../lib/Matrix_Factorization.R")
```


#### Step 2.2 Parameter Tuning
Here you should tune parameters, such as the dimension of factor and the penalty parameter $\lambda$ by cross-validation.
```{r}
source("../lib/cross_validation.R")
f_list <- seq(10, 20, 10)
l_list <- seq(-2, -1, 1)
f_l <- expand.grid(f_list, l_list)
```

```{r, eval=FALSE}
result_summary <- array(NA, dim = c(nrow(f_l), 100, 4)) 
run_time <- system.time(for(i in 1:nrow(f_l)){
    par <- paste("f = ", f_l[i,1], ", lambda = ", 10^f_l[i,2])
    cat(par, "\n")
    current_result <- cv.function.als(data, K = 5, f = f_l[i,1], lambda = 10^f_l[i,2])
    result_summary[,,i] <- matrix(unlist(current_result), ncol = 100, byrow = T) 
    save(result_summary, file = "../output/rmse_als.Rdata")
  
})
```

```{r}
load("../output/rmse_als.Rdata")
rmse <- data.frame(rbind(t(result_summary[1,,]), t(result_summary[2,,])), train_test = rep(c("Train", "Test"), each = 4), par = rep(paste("f = ", f_l[,1], ", lambda = ", 10^f_l[,2]), times = 2)) %>% gather("epoch", "RMSE", -train_test, -par)
rmse$epoch <- as.numeric(gsub("X", "", rmse$epoch))
rmse %>% drop_na(RMSE) %>% ggplot(aes(x = epoch, y = RMSE, col = train_test)) + geom_point() + facet_grid(~par)
```
```{r}
test_rmse <- rep(NA, 4)
for (i in 1:nrow(f_l)) {
  epoch_rmse <- result_summary[2,,i]
  epoch_rmse <- epoch_rmse[!is.na(epoch_rmse)]
  test_rmse[i] <- epoch_rmse[length(epoch_rmse)]
}
best_f <- f_l[which(test_rmse==min(test_rmse)), 1]
best_lambda <- 10**f_l[which(test_rmse==min(test_rmse)), 2]
```

#### Step 2.3 Evaluation on the Model without Postprocessing
```{r, eval= FALSE}
result <- als(f = best_f, lambda = best_lambda, max.iter = 100, stopping.thres = 0.001,
                   data = data, train = data_train, test = data_test)

save(result, file = "../output/mat_fac_als.RData")
```

You should visualize training and testing RMSE by different epochs ([One Epoch is when an ENTIRE dataset is passed forward and backward through the neural network only ONCE](https://towardsdatascience.com/epoch-vs-iterations-vs-batch-size-4dfb9c7ce9c9)). 

```{r}
load(file = "../output/mat_fac_als.RData")
library(ggplot2)

RMSE <- data.frame(epochs = seq(1, 100, 1), Training_MSE = result$train_RMSE, Test_MSE = result$test_RMSE) %>% gather(key = train_or_test, value = RMSE, -epochs)

RMSE %>% drop_na(RMSE) %>% ggplot(aes(x = epochs, y = RMSE,col = train_or_test)) + geom_point()
```

### Step 3 Postprocessing
After matrix factorization, postporcessing will be performed to improve accuracy.
The referenced papers are:

P1:[Global bias correction](./paper/P4 Large-scale Parallel Collaborative Filtering for the Netflix Prize.pdf) Section 4.1

P2:[Postprocessing SVD with KNN](./paper/P2 Improving regularized singular value decomposition for collaborative filtering .pdf) Section 3.5

P3:[Postprocessing SVD with kernel ridge regression](./paper/P2 Improving regularized singular value decomposition for collaborative filtering .pdf) Section 3.6

P4:[Linearly combination of predictors](./paper/P4 Large-scale Parallel Collaborative Filtering for the Netflix Prize.pdf) Section 4.1


```{r}
pred_rating <- t(result$q) %*% result$p
#define a function to extract the corresponding predictedrating for the test set.
extract_pred_rating <- function(test_set, pred){
  pred_rating <- pred[as.character(test_set[2]), as.character(test_set[1])]
  return(pred_rating)
}
#extract predicted rating
pred_test_rating <- apply(data_test, 1, extract_pred_rating, pred_rating)

#mean(P)
pred_mean <- mean(pred_test_rating)
#mean(test)
mean_test_rating <- mean(data_test$rating)

#mean(test) - mean(P)
mean_diff <- mean_test_rating - pred_mean

data_test$pred <- pred_test_rating
data_test$pred_adj <- pred_test_rating + mean_diff

boxplot(data_test$pred_adj ~ data_test$rating)
#calculate RMSE
rmse_adj <- sqrt(mean((data_test$rating - data_test$pred_adj)^2))
cat("The RMSE of the adjusted model is", rmse_adj)
```

Depending on your postprocessing method, you might want to cross-validate on some parameters related to the postprocessing. Don't forget to visualize the cross-validation process through graphs.


#### Step 3.2 P2 Postprocessing SVD with KNN
```{r}




```


#### Step 3.3 P3 Postprocessing SVD with kernel ridge regression


```{r}
# frequent movie
library(tidyverse)
topnum=500
top=data%>%
  mutate(movieId=as.character(movieId))%>%
  count(movieId)%>%
  arrange(desc(n))%>%
  head(topnum)%>%
  select(movieId)%>%
  pull(movieId)
top

## # of movies rated by each user in train dataset

num_of_rated=data_train%>%
  count(userId)%>%
  pull(n)


select_movie=function(i){
  if (num_of_rated[i]<topnum) {
    movie=data_train%>%
      filter(userId==i)%>%
      pull(movieId)
    movie_in_top=movie[movie %in% top]
    addition=top[!top %in% movie_in_top][1:(topnum-length(movie))]
    
    sel_list=c(movie,addition)
  }
  else
  {
    movie=data_train%>%
      filter(userId==i)%>%
      pull(movieId)
    sel_list=movie[1:topnum]
  }
  return(as.character(sel_list))
}


```


## new version
```{r}
krr.start=proc.time()

X=scale(result$q,center = F)
X=t(X)
lambda=2 #tuning parameter
user=as.character(1:U)
KRR_pred=matrix(0,length(unique(data$movieId)),length(user))
#str(KRR_pred) # 9724* 610

for (i in 1:length(user)){
  topmovie=select_movie(i)
  topmovie
  y=pred_rating[topmovie,user[i]] # y--topnum*1 user i's rating for top topnum movie
  y
  # X-- factors for top topnum movies topnum*10
  Xtop=X[topmovie,]
  Xtop
  K=exp(2*(Xtop%*%t(Xtop)-1)) # topnum*topnum
  left=exp(2*(X%*%t(Xtop)-1))
  b=solve(K+lambda*diag(nrow(K)))
  a=b%*%y #topnum*1
  # str(beta)
  # str(t(Xtop))
  KRR_pred[,i]=left%*%a
  
}

rownames(KRR_pred)=levels(as.factor(data$movieId))
colnames(KRR_pred)=as.character(1:U)
KRR_test <- apply(data_test, 1, extract_pred_rating, KRR_pred)
data_test$KRR <- KRR_test

rmse_KRR <- sqrt(mean((data_test$rating - data_test$KRR)^2))
rmse_KRR

krr.stop=proc.time()
timeused=krr.stop-krr.start
timeused
```


## last version
```{r}

slow.start=proc.time()

## slow 

X=scale(result$q,center=F)

# str(X)

# rownames(X)

X=t(X)

Xtop=X[top,]

lambda=1 #tuning parameter

user=as.character(1:U)

KRR_pred=matrix(0,length(unique(data$movieId)),length(user))

#str(KRR_pred) # 9724* 610



K=exp(2*(Xtop%*%t(Xtop)-1)) # topnum*topnum

left=exp(2*(X%*%t(Xtop)-1))

b=solve(K+lambda*diag(nrow(K)))



for (i in 1:length(user)){

  y=pred_rating[top,user[i]] # y--topnum*1 user i's rating for top topnum movie

  # X-- factors for top topnum movies topnum*10

  # xi-- specific movie's factor

  

  a=b%*%y #topnum*1

  

  # str(beta)

  # str(t(Xtop))

  

  KRR_pred[,i]=left%*%a

  

}



rownames(KRR_pred)=levels(as.factor(data$movieId))

colnames(KRR_pred)=as.character(1:U)

KRR_test <- apply(data_test, 1, extract_pred_rating, KRR_pred)

data_test$KRR <- KRR_test



rmse_KRR <- sqrt(mean((data_test$rating - data_test$KRR)^2))

rmse_KRR



slow.stop=proc.time()

timeused=slow.stop-slow.start

timeused
```